---
title: "Antibiotics"
author: "Anna Letcher Hartman"
date: 2023
output:
  github_document
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Creating effective data visualizations is an *iterative*
process; very rarely will the first graph you make be the most
effective. The most effective thing you can do to be successful in this
iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely
dependent on *the question you are trying to answer*. A visual that is
totally ineffective for one question may be perfect for answering a
different question.

In this challenge, you will practice *iterating* on data visualization,
and will anchor the *assessment* of your visuals using two different
questions.

*Note*: Please complete your initial visual design **alone**. Work on
both of your graphs alone, and save a version to your repo *before*
coming together with your team. This way you can all bring a diversity
of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the
[*minimum inhibitory
concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration)
(MIC) of three drugs for different bacteria. The smaller the MIC for a
given drug and bacteria pair, the more practical the drug is for
treating that particular bacteria. An MIC value of *at most* 0.1 is
considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin,
streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized
into a genus based on a number of features, including their resistance
to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the
data** `df_antibiotics`. These **cannot** be simple variations on the
same graph; for instance, if two of your visuals could be made identical
by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16
bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of
ideas, and don't fixate on any one idea for too long. You will have a
chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be
helpful to `pivot_longer()` the data to make certain visuals easier to
construct.

```{r}
theme_common <- function() {
  theme_minimal() %+replace%
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(4, 4, 4, 4), size = 16),
    axis.title.y = element_text(margin = margin(4, 4, 4, 4), size = 16, angle = 90),

    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),

    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 12),

    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_line(color = "grey90"),

    aspect.ratio = 4 / 4,

    plot.margin = unit(c(t = +0, b = +0, r = +0, l = +0), "cm"),
    plot.title = element_text(size = 18),
    plot.title.position = "plot",
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 12)
  )
}

df_antibiotic_long <- 
  df_antibiotics %>%
  pivot_longer(
    names_to = "antibiotic",
    values_to = "mic",
    cols = - c(bacteria, gram)
  )

df_antibiotic_long
```

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE

df_antibiotic_long %>%
  #mutate(logMic = log(mic)) %>% 
  arrange(gram) %>% 
  ggplot(aes(antibiotic, bacteria, fill = mic)) +
  geom_tile() +
  scale_fill_continuous(trans = "log10") +
  facet_wrap( ~gram)

```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.2}
 
df_antibiotic_long %>%
  ggplot(aes(mic, bacteria, fill = antibiotic)) +
  geom_col(position = "dodge")+
  facet_wrap( ~gram) +
  scale_x_log10()


```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE
df_antibiotic_long %>%
  mutate(HumanuseMic = na_if(mic, mic <= 0.1)) %>% 
  #filter(mic <= 0.1) %>% 
  ggplot(aes(bacteria, HumanuseMic, fill = antibiotic)) + #, alpha = gram
  geom_col(position = "dodge") +
  #scale_alpha_discrete(range = c(0.4, 1)) +
  coord_flip() +
  facet_wrap( ~gram) +
  #theme_common()+
  ylim(0, 0.1)
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE
wid <- 0.5

df_antibiotic_long %>%
  #arrange(desc(mic)) %>% 
  ggplot(aes(mic, bacteria, color = antibiotic, shape = gram)) +
  geom_point(position = position_dodge(width = wid)) +
  #theme_common()+
  xlim(0, 0.1)
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE
df_antibiotics

df_antibiotics %>%
  ggplot(aes(neomycin, penicillin, color = bacteria, shape = gram)) +
  geom_point() + 
  # geom_text(
  #    aes(label = bacteria),
  #    
  #    nudge_x = 0.05,
  #    #hjust = 0, 
  #    nudge_y = 0.5
  #   # label = 'bacteria',
  #   #check_overlap = TRUE
  #   ) +
  # geom_label(
  #   aes(label = bacteria), 
  #   # position = "doge",
  #   nudge_x = .5,
  #   nudge_y = .5,
  #   size = 3,
  #   label.padding = unit(0, "lines"),
  #   label.r = unit(0, "lines"),
  #   label.size = 0,
  #   show.legend = FALSE
  # #            ) +
  # theme(legend.position = "none")+
  scale_x_log10() +
  scale_y_log10()
```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer
both Guiding Questions. Note that you must also identify which of your
five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in
answering the questions below. You may need to revise one or more of
your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most
effective at helping answer both guiding questions. **Use this as an
opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against
> bacteria of different genera and Gram stain? penicillin, streptomycin,
> and neomycin

*Observations* - What is your response to the question above? -
streptomycin is effective in treating humans against only 4 bacteria in
this data set 3 of witch are gram positive and 1 gram negative. Outside
of treating humans it has a mid range effectiveness. Neomycin is
effective in treating humans against 9 bacteria in this data set, 5 gram
negative and 4 gram positive, the largest number of any antibiotic.
Penicillin had the widest spread of effectiveness with the highest and
lowest mic. It is effective in treating humans against 6 gram positive
bacteria in this data set.

\- Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question? - 2 and 4

-   Why? - two helped me see the bigger picture and four made it easy to
    see how effective they actually were in humans since I limited it to
    showmic only mic values at or under the effectiveness threshold for
    humans

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*,
and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*
[2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus
> pneumoniae*?

*Observations* - What is your response to the question above? - In
figure five you can see that Diplococcus pneumoniae is clustered with
all of the *Streptococcus* bacteria, except Streptococcus fecalis,
having a low mic for penicilin and a high mic for neomicin. Where as
Streptococcus fecalis is closer to a cluster of gram negative bacteria
with mid range mic for both antibiotics.

\- Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question? - 5

\- Why? - five very clearly shows how most similar bacteria have similar
reactions to penicillin and neomicin with the notable exception of these
two bacteria

References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide
antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009)
[link](https://www.americanscientist.org/article/thats-funny)
